---
title: "Data Analysis Report - Predicting Animal Adoption at Long Beach Animal Shelter"
author: "Arjun Narang, Lexie Wheeler, Adithi Math, Kimball Otteson"
format:
  pdf:
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 12pt
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

## Good practice to include libraries and functions you will need in a chunk
## at the beginning of your file
library(jtools)
library(patchwork)
library(tidyverse)
## functions
logit <- function(x) log(x/(1 - x))
invlogit <- function(x) exp(x)/(1 + exp(x))

## This sets the gpplot theme to white for all plots
## the option `base_size` sets the base font size for all plots
theme_set(
    theme_bw(base_size = 8)
)
```

# Abstract

This project investigates the factors that influence pet adoption at the Long Beach Animal Shelter, using a dataset provided via the {animalshelter} R package and TidyTuesday repository. With over 29,000 animal records, we aim to identify the variables that significantly impact the likelihood of an animal being adopted.

Our research focuses on predicting adoption (yes/no) using a logistic regression model. Key predictors include species animal_type, intake condition, sex, intake type, primary color, and season of outcome. After cleaning and transforming the dataset—including recoding the outcome into a binary variable and engineering a season variable—we explore adoption rates across each predictor and fit a Bernoulli logistic model to estimate the probability of adoption.

# Introduction

In this section, introduce the problem and question(s) of interest that you aim to address in your analysis. Provide relevant background information and explain the motivation and significance of your research question. Cite manuscripts or websites you use. Here an example of how to cite a paper, @wickham2019. See  @quarto_cit for how to manage citations in Quarto.

Animal shelters play a vital role in animal welfare, often facing resource constraints while handling high intake volumes. One pressing issue is determining which animals are most likely to be adopted and how different factors affect this likelihood. Better understanding of adoption patterns can inform targeted interventions to improve outcomes.

Our main research questions are:

Are certain types of pets more likely to be adopted than others?

How do factors like intake condition, sex, intake type, and season influence adoption likelihood?

Using data from the Long Beach Animal Shelter, we aim to model adoption probabilities and highlight actionable trends. This work builds on the goal of improving animal outcomes through evidence-based strategies.


# Data and Methods

Describe the data you used in your analysis, including the source, characteristics, and any pre-processing or cleaning steps you performed. 



```{r}
# Load Data
longbeach_partial <- readRDS("data/longbeach_partial.rds")

# Create a binary variable: 1 if adopted, 0 otherwise
longbeach <- longbeach_partial %>%
  mutate(adopted = ifelse(tolower(trimws(outcome_type)) == "adoption", 1, 0))

# Make animal type a factor
longbeach <- longbeach %>%
  mutate(
    animal_type = as.factor(animal_type),
    intake_condition = as.factor(intake_condition),
    sex = as.factor(sex),
    intake_type = as.factor(intake_type)
  )

# Make seasons
longbeach <- longbeach %>%
  mutate(season = case_when(
    format(outcome_date, "%m") %in% c("12", "01", "02") ~ "Winter",
    format(outcome_date, "%m") %in% c("03", "04", "05") ~ "Spring",
    format(outcome_date, "%m") %in% c("06", "07", "08") ~ "Summer",
    format(outcome_date, "%m") %in% c("09", "10", "11") ~ "Fall",
    TRUE ~ NA_character_
  ))
longbeach_clean <- na.omit(longbeach %>%
                             select(adopted, animal_type, intake_condition, sex, intake_type, outcome_date, primary_color, season))

# Make larger categories of colors

longbeach <- longbeach %>%
  mutate(color = case_when(
    tolower(primary_color) %in% c("tricolor", "calico", "tabby", "ticked", "dapple", "blue merle", "blue tick", "brown tiger", "liver tick", "point", "seal point") ~ "Patterns",
    tolower(primary_color) %in% c("brown", "brown tabby", "brown brindle", "ch lynx point", "chocolate", "chocolate point", "brown merle", "sable", "snowshoe", "tan", "yellow brindle") ~ "Browns",
    tolower(primary_color) %in% c("blue", "blue brindle", "blue lynx point", "blue cream", "blue point", "blue tabby", "gray", "gray tabby", "gray tiger", "lynx point", "seal", "silver", "silver tabby", "st lynx point", "tortie dilute") ~ "Grays and blues", 
    tolower(primary_color) %in% c("apricot", "calico dilute", "calico point", "calico tabby", "fawn", "lilac_cream point", "liver", "orange", "orange tabby", "peach", "pink", "red", "red merle", "red point", "ruddy") ~ "Orange and Red",
    tolower(primary_color) %in% c("black", "black lynx point", "black smoke", "black tabby", "black tiger", "torbie", "tortie") ~ "Black or Dark",
    tolower(primary_color) %in% c("white", "flame point", "lilac lynx point", "lilac point", "silver lynx point") ~ "White Based",
    tolower(primary_color) %in% c("green") ~ "Greens",
    tolower(primary_color) %in% c("yellow", "blonde", "buff", "cream", "cream point", "cream tabby", "gold", "wheat") ~ "Cream and yellows",
    TRUE ~ NA_character_
  ))



numberAdopted <- 6290/(23310+6290) *100

# Count and proportion of adopted vs. non-adopted animals
adoption_counts <- longbeach %>%
  mutate(adoption_status = ifelse(adopted == 1, "Adopted", "Not Adopted")) %>%
  count(adoption_status) %>%
  mutate(proportion = n / sum(n))


# Check Adoption Rates
adoption_rates <- longbeach_clean %>%
  group_by(animal_type) %>%
  summarise(adoption_rate = mean(adopted) * 100) %>%
  arrange(desc(adoption_rate))

print(adoption_rates)


# Number of animals per animal type
animal_counts <- longbeach %>%
  count(animal_type)
```




```{r}
# Number of animals per animal type
animal_counts <- longbeach %>%
  count(animal_type)

# Plotting raw numbers to just see how many of each animal we have
animals <- ggplot(animal_counts, aes(x = reorder(animal_type, -n), y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Number of Animals per Animal Type",
       x = "Animal Type",
       y = "Number of Animals")

# Adoption Rates by Animal Type
rateByType <- ggplot(adoption_rates, aes(x = reorder(animal_type, -adoption_rate), y = adoption_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Adoption Rates by Animal Type",
       x = "Animal Type",
       y = "Adoption Rate (%)") +
  theme_minimal()

# Plot proportions
adoptionProp <- ggplot(adoption_counts, aes(x = adoption_status, y = proportion, fill = adoption_status)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::percent_format()) +  # Convert to percentage format
  labs(title = "Proportion of Adopted vs. Non-Adopted Animals",
       x = "Adoption Status",
       y = "Proportion (%)") +
  theme_minimal()

longbeach <- longbeach %>%
  mutate(outcome_date = as.Date(outcome_date))


seasonal_adoption <- longbeach %>%
  filter(!is.na(season) & !is.na(adopted)) %>%  # Remove NAs
  group_by(season) %>%
  summarise(adoption_rate = mean(adopted, na.rm = TRUE) * 100) %>%
  arrange(factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))  # Ensure correct order

print(seasonal_adoption)

seasonalPlot <- ggplot(seasonal_adoption, aes(x = season, y = adoption_rate, fill = season)) +
  geom_bar(stat = "identity") +
  labs(title = "Seasonal Trends in Pet Adoptions",
       x = "Season",
       y = "Adoption Rate (%)") +
  theme_minimal()


# Adoption rate by intake condition
intakeCondition <- longbeach %>%
  filter(!is.na(intake_condition) & !is.na(adopted)) %>%
  group_by(intake_condition) %>%
  summarise(adoption_rate = mean(adopted) * 100) %>%
  arrange(desc(adoption_rate))

# Plot adoption rate by intake condition
rateByCondition <- ggplot(intakeCondition, aes(x = reorder(intake_condition, -adoption_rate), y = adoption_rate)) +
  geom_col(fill = "steelblue") +
  labs(title = "Adoption Rates by Intake Condition",
       x = "Intake Condition",
       y = "Adoption Rate (%)") +
  theme_minimal() + coord_flip()


# Adoption rate by Sex
rateSex <- longbeach %>%
  filter(!is.na(sex) & !is.na(adopted)) %>%
  group_by(sex) %>%
  summarise(adoption_rate = mean(adopted) * 100) %>%
  arrange(desc(adoption_rate))

# Plot adoption rate by Sex
rateBySex <- ggplot(rateSex, aes(x = reorder(sex, -adoption_rate), y = adoption_rate)) +
  geom_col(fill = "steelblue") +
  labs(title = "Adoption Rates by Sex",
       x = "Sex",
       y = "Adoption Rate (%)") +
  theme_minimal()

# Adoption rate by intake type - different than intake condition
# There are 2 different types of returns so I combined them here using mutate
intakeType <- longbeach %>%
  filter(!is.na(intake_type) & !is.na(adopted)) %>%
  mutate(intake_type = case_when(
    intake_type %in% c("return", "adopted animal return") ~ "return",
    TRUE ~ intake_type)) %>%
  group_by(intake_type) %>%
  summarise(adoption_rate = mean(adopted) * 100) %>%
  arrange(desc(adoption_rate))

# Plot adoption rate by intake type
rateByType <- ggplot(intakeType, aes(x = reorder(intake_type, -adoption_rate), y = adoption_rate)) +
  geom_col(fill = "steelblue") +
  labs(title = "Adoption Rates by Intake Type",
       x = "Intake Type",
       y = "Adoption Rate (%)") +
  theme_minimal() + coord_flip() 

longbeach <- longbeach %>%
  mutate(color = case_when(
    tolower(primary_color) %in% c("tricolor", "calico", "tabby", "ticked", "dapple", "blue merle", "blue tick", "brown tiger", "liver tick", "point", "seal point") ~ "Patterns",
    tolower(primary_color) %in% c("brown", "brown tabby", "brown brindle", "ch lynx point", "chocolate", "chocolate point", "brown merle", "sable", "snowshoe", "tan", "yellow brindle") ~ "Browns",
    tolower(primary_color) %in% c("blue", "blue brindle", "blue lynx point", "blue cream", "blue point", "blue tabby", "gray", "gray tabby", "gray tiger", "lynx point", "seal", "silver", "silver tabby", "st lynx point", "tortie dilute") ~ "Grays and blues", 
    tolower(primary_color) %in% c("apricot", "calico dilute", "calico point", "calico tabby", "fawn", "lilac_cream point", "liver", "orange", "orange tabby", "peach", "pink", "red", "red merle", "red point", "ruddy") ~ "Orange and Red",
    tolower(primary_color) %in% c("black", "black lynx point", "black smoke", "black tabby", "black tiger", "torbie", "tortie") ~ "Black or Dark",
    tolower(primary_color) %in% c("white", "flame point", "lilac lynx point", "lilac point", "silver lynx point") ~ "White Based",
    tolower(primary_color) %in% c("green") ~ "Greens",
    tolower(primary_color) %in% c("yellow", "blonde", "buff", "cream", "cream point", "cream tabby", "gold", "wheat") ~ "Cream and yellows",
    TRUE ~ NA_character_
  ))

# Adoption rate by Primary Color
rateColor <- longbeach %>%
  filter(!is.na(color) & !is.na(adopted)) %>%
  group_by(color) %>%
  summarise(adoption_rate = mean(adopted) * 100) %>%
  arrange(desc(adoption_rate))

# Plot Adoption Rate by colors using only the top 20 with the highest adoption
rateByColor <- ggplot(rateColor, aes(x = reorder(color, -adoption_rate), y = adoption_rate, fill = color)) +
  geom_col() +
  labs(title = "Adoption Rates by Primary Color",
       x = "Primary Color",
       y = "Adoption Rate (%)") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c(
    "Patterns" = "darkorchid",
    "Browns" = "brown",
    "Grays and blues" = "steelblue",
    "Orange and Red" = "red",
    "Black or Dark" = "black",
    "White Based" = "lightgray",
    "Greens" = "darkgreen",
    "Cream and yellows" = "wheat"
  ))


combined <- rateByColor + rateByCondition +rateBySex + rateByType
combined1 <-  seasonalPlot + adoptionProp 
combined2 <- animals


```

```{r}
#| label: fig-eda
#| fig-cap: "These four figures show the plots that are shown with respect to adoption rates. The titles explain what we are looking at in these graphs. Based on this priliminary analysis, we were able to guide our thinking and questions with more knowledge and understanding of the data. "
#| fig-width: 10
#| fig-height: 7
#| fig-show: hold

# Display the combined plot
combined

#| label: fig-eda1
#| fig-cap: "Here are two plots showing first the adoption rates per season. We see that Spring has the lowest adoption rate as Winter has the highest. The second plot on the right is simply the adopted rate vs. non adopted. Not adopted is categorized as anything but adopted. We see that there is a higher proportion for non-adopted "
#| fig-width: 5
#| fig-height: 10
#| fig-show: hold

combined1

#| label: fig-eda1
#| fig-cap: "This plot is a little different than the rest as it is not plotted vs proportion. It is simply just the raw numbers of each kind of animal we have. This helped us to better understand what we were working with as it is easy to assume that just cat and dogs are being adopted. It also helps us to understand that an animal could have a large adoption rate because there are not many types of that animal."
#| fig-width: 5
#| fig-height: 10
#| fig-show: hold

combined2

```


## Data Description and Preprocessing

The dataset for this project comes from the Long Beach Animal Shelter, obtained through the `animalshelter` R package and the TidyTuesday project (March 4, 2025). It contains over 29,000 records on individual animals, including species, intake condition, intake type, sex, primary color, outcome type, and outcome date.

### Data Cleaning Steps

To prepare the data for analysis, we performed the following steps:

- **Outcome Recoding**:  
  We created a binary variable `adopted`, where:
  - `1` = the animal was adopted (outcome_type == "Adoption")
  - `0` = all other outcome types (e.g., transfer, return, euthanasia)

- **Factor Conversion**:  
  Categorical variables were converted into factors for modeling:
  - `animal_type`, `intake_condition`, `sex`, `intake_type`, and `primary_color`

- **Season Variable Engineering**:  
  We created a `season` variable from `outcome_date`:
  - Winter: December–February  
  - Spring: March–May  
  - Summer: June–August  
  - Fall: September–November

- **Color Grouping**:  
  The original `primary_color` variable contained many highly specific labels (e.g., “blue merle”, “calico dilute”). We grouped these into broader, interpretable categories stored in a new `color` variable:
  - *Patterns*: e.g., calico, tabby, merle
  - *Browns*: e.g., brown, chocolate, sable
  - *Grays and blues*: e.g., blue, gray, silver
  - *Orange and red*: e.g., red, orange, ruddy
  - *Black or dark*: e.g., black, tortie
  - *White based*: e.g., white, lilac point
  - *Cream and yellows*: e.g., cream, gold, wheat
  - *Greens*: retained as a rare but existing entry

- **Missing Data**:  
  To ensure model integrity, we excluded rows with missing values in any of the analysis variables using `na.omit()`.

The resulting cleaned dataset, named `longbeach_clean`, includes the following variables:
- `adopted`
- `animal_type`
- `intake_condition`
- `sex`
- `intake_type`
- `outcome_date`
- `primary_color`
- `season`
- `color`

This version was saved as an `.rds` file for reproducibility and used throughout the rest of the analysis.





This is a good place where to include results of your exploratory data analysis. See @fig-eda as an example. 
Check the .qmd file for an example of how to include plots, control their size and include references in the text. 


In our exploratory data analysis, we found that overall, there was a higher proportion of animals that were not adopted compared to adopted. Not adopted means everything other than adopted. So the animal could have passed away or been sent back, just not adopted. We also found that most of the plots created vs adoption rates have some categories that stick out more than the rest. The only plots that did not were color and seasons which were two categories that we mutated ourselves. All of the plots created during our exploratory data analysis can be seen above with comments to explain the graphs and how they helped us come to our overall conclusion.



#You should also explain the methods you employed. I expect you to define the statistical model you chose for #your data in mathematical form, as we did throughout the course. 

#Crab example: 

We define as outcome the adoption for each animal and denote the observed values as $y_i$ with $i = 1, \ldots, n$ with $n = 29600$. Let $y_i = 1$ indicate an animal being adopted and $0$ otherwise. 

We assume that each $y_i$ is a realization of a random variables $Y_i \sim \mbox{Bernoulli}(p_i)$ independently, and model the probability of an animal being adopted using the following logistic regression:

$$
\begin{aligned}
  \mathrm{logit}(p_i) =\ & \beta_0 
  + \beta_1 \cdot \mathrm{I}(\text{animal\_type}_i = \text{amphibian}) 
  + \beta_2 \cdot \mathrm{I}(\text{animal\_type}_i = \text{bird}) //
  + \beta_3 \cdot \mathrm{I}(\text{animal\_type}_i = \text{cat}) 
  + \beta_4 \cdot \mathrm{I}(\text{animal\_type}_i = \text{guinea pig}) 
  + \beta_5 \cdot \mathrm{I}(\text{animal\_type}_i = \text{livestock}) \\
  + \beta_6 \cdot \mathrm{I}(\text{animal\_type}_i = \text{other}) 
  + \beta_7 \cdot \mathrm{I}(\text{animal\_type}_i = \text{rabbit}) 
  + \beta_8 \cdot \mathrm{I}(\text{animal\_type}_i = \text{reptile}) \\
  + \beta_9 \cdot \mathrm{I}(\text{animal\_type}_i = \text{wild}) 
  + \beta_{10} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{aged}) 
  + \beta_{11} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{behavior mild})\\ 
  + \beta_{12} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{behavior moderate}) 
  + \beta_{13} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{behavior severe}) 
  + \beta_{14} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{feral}) \\
  + \beta_{15} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{fractious}) 
  + \beta_{16} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{i/i report}) 
  + \beta_{17} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{ill mild}) \\
  + \beta_{18} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{ill moderate}) 
  + \beta_{19} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{ill severe}) 
  + \beta_{20} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{injured mild}) \\ 
  + \beta_{21} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{injured moderate}) 
  + \beta_{22} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{injured severe}) 
  + \beta_{23} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{intakeexam}) \\
  + \beta_{24} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{under age/weight}) 
  + \beta_{25} \cdot \mathrm{I}(\text{intake\_condition}_i = \text{welfare seizures}) \\
  + \beta_{26} \cdot \mathrm{I}(\text{sex}_i = \text{Female}) 
  + \beta_{27} \cdot \mathrm{I}(\text{sex}_i = \text{Neutered}) \\
  + \beta_{28} \cdot \mathrm{I}(\text{sex}_i = \text{Spayed}) 
  + \beta_{29} \cdot \mathrm{I}(\text{intake\_type}_i = \text{confiscate}) \\
  + \beta_{30} \cdot \mathrm{I}(\text{intake\_type}_i = \text{euthanasia required}) 
  + \beta_{31} \cdot \mathrm{I}(\text{intake\_type}_i = \text{foster}) \\
  + \beta_{32} \cdot \mathrm{I}(\text{intake\_type}_i = \text{owner surrender}) 
  + \beta_{33} \cdot \mathrm{I}(\text{intake\_type}_i = \text{quarantine}) 
  + \beta_{34} \cdot \mathrm{I}(\text{intake\_type}_i = \text{return}) \\
  + \beta_{35} \cdot \mathrm{I}(\text{intake\_type}_i = \text{safe keep}) 
  + \beta_{36} \cdot \mathrm{I}(\text{intake\_type}_i = \text{stray}) \\
  + \beta_{37} \cdot \mathrm{I}(\text{intake\_type}_i = \text{trap, neuter, return}) 
  + \beta_{38} \cdot \mathrm{I}(\text{intake\_type}_i = \text{welfare seized}) 
  + \beta_{39} \cdot \mathrm{I}(\text{intake\_type}_i = \text{wildlife}) \\
\end{aligned}
$$



where $\mbox{I}()$ indicates a dummy variable, taking value 1 when the condition is true and zero otherwise. 

```{r}

m1 = glm(adopted ~ animal_type, data = longbeach_clean)
nyew2 = data.frame(animal_type = levels(longbeach_clean$animal_type))
pred2 = predict(m1, newdata = nyew2, type = "link", se.fit = T)

nyew2$prob = invlogit(pred2$fit)

nyew2$prob_lower = invlogit(pred2$fit - 1.96*pred2$se.fit)
nyew2$prob_upper = invlogit(pred2$fit + 1.96*pred2$se.fit)

anyimal <- ggplot(longbeach_clean, aes(x = animal_type, y = adopted)) +
  geom_jitter(height = 1/80, size = 3, alpha = 1/4) +
  geom_pointrange(aes(y = prob, ymin = prob_lower, ymax = prob_upper), 
                  data = nyew2, size = 1, color = scales::muted('blue')) +
  labs(title = "Probability of Adoption based on Animal Type", x = "Animal Type", y = "p(Adopted)")
print(anyimal)

gm42 = glm(adopted ~ animal_type + intake_condition + sex + intake_type, data = longbeach_clean, family = binomial)

resid.df = data.frame(fitted = gm42$fitted.values,
                      deviance = resid(gm42, type = "deviance"), 
                      Adopted = longbeach_clean$adopted,
                      Animal = longbeach_clean$animal_type,
                      intake_condition = longbeach_clean$intake_condition,
                      Sex = longbeach_clean$sex,
                      intake_type = longbeach_clean$intake_type)

p_resid = ggplot(resid.df, aes(x = fitted, y = deviance)) +
  geom_point() +
  geom_hline(yintercept = 0, col = 2) +
  labs(title = "Residuals", x = "Fitted", y = "Deviance")
print(p_resid)
```

# Results

The final logistic regression model included four key predictors: animal_type, intake_condition, sex, and intake_type. The model with these variables achieved the lowest AIC, indicating the best fit among the tested models. The baseline animal for comparison is a normally behaved, male, foster-intake dog.

## Animal Type
Adoption likelihood varied significantly by species:
- Birds were 2.84 times more likely to be adopted than the baseline.
- Cats and rabbits also had significantly higher odds of adoption, with log-odds of 1.15 and 0.56.
- Livestock had the highest effect, with a log-odds coefficient of 13.11, though this may reflect a small sample size or special intake circumstances.
- Other animals like reptiles were 1.80 log-odds more likely to be adopted than the baseline dog.

## Intake Condition

Animals arriving in better condition were more likely to be adopted:
- Mildly ill or injured animals showed higher adoption likelihoods than those with severe conditions. For example:
    - Ill (mild): 1.13
    - Injured (mild): 0.88
    - Under age/weight animals were also more likely to be adopted (1.25).
- Behavioral concerns reduced adoption probability:
    - Severe behavior issues: 0.30
    - Feral: 0.07
    - Fractious (aggressive): 0.18

## Sex
Sterilization status had a strong impact:
- Neutered animals: log-odds = 11.10
- Spayed animals: log-odds = 11.36
- This shows that sterilized animals were far more likely to be adopted than intact males (the baseline group).

## Intake Type
- Animals coming in through Trap-Neuter-Return (TNR) programs had much lower chances of adoption, with a coefficient of 0.03, indicating that these animals were likely returned to their environment rather than adopted.
- Foster intake (baseline) served as the reference for comparison.


## Now looking at the peason residual models

Model Diagnostics: Pearson Residuals
To assess model fit, we examined Pearson residuals versus fitted values for each predictor included in the final logistic regression model.


1. Animal Type
- Most residuals are clustered around 0, indicating that the model fits well for most categories of animal type.
- A few points with residuals above 1 suggest some minor underfitting (for instance, maybe actual adoption outcomes were higher than predicted) for certain less common animal types.
- However, no residuals exceed ±2, indicating that there are no extreme outliers or poor-fitting categories.
Conclusion: The model captures adoption patterns by animal type reasonably well.

2. Intake Condition
- Once again, the residuals are tightly clustered around zero.
- Similar to animal type, mild under- and over-prediction appears for certain conditions, especially at lower fitted probabilities.
- No concerning deviations or extreme values are observed.
Conclusion: The intake condition variable is well-modeled, with no evidence of poor fit for any specific category.

3. Sex

- The residuals are mostly centered around zero, with fewer distinct fitted values due to the small number of sex categories.
- Slight underfitting is seen in some categories, possibly due to class imbalance (like fewer unknown animals).
Conclusion: The sex variable contributes significantly to prediction, and residual patterns are acceptable.

4. Intake Type

- There is a wider spread of residuals, especially for low fitted values.
- This is most likely due to variation in adoption rates between standard intake types (like “foster” vs. “TNR”) and rarer or edge-case types.
Conclusion: The model performs well overall but could benefit from further refinement or splitting of nuanced intake categories.

Overall Model Fit
Across all predictors:
- Most residuals fall within the ±2 range, indicating no strong violations of model assumptions.
- The horizontal spread of residuals proves there is no clear pattern of heteroscedasticity.
- These plots all collectively support the use of the chosen predictors, and the model appears to generalize well to the observed data.

## Looking at AIC output and Stepwise Regression

Step 1: Starting from the Null Model

Null model AIC = 30623.21
- Adding sex reduces the AIC to 22953.39, a substantial improvement of nearly 7,670 points, making it the strongest single predictor.
- Other predictors also lowered the AIC:
  - intake_type: AIC = 25533.36
  - animal_type: AIC = 26776.34
  - intake_condition: AIC = 29932.68

Step 2: Adding to the Model with sex
The updated model is adopted ~ sex (AIC = 22953.39)
- Adding intake_condition yields a large drop in AIC to 22216.39, proving this variable meaningfully improves model performance
- Adding other variables instead gives higher AICs:
  - intake_type: AIC = 22616.49
  - animal_type: AIC = 22769.29

Step 3: Adding to the Model with sex + intake_condition
Current AIC = 22216.39
- Adding intake_type lowers AIC further to 21928.34
- Other options:
  - animal_type: AIC = 22021.64
  - intake_type is selected as the next best variable.
Summary of Optimal Model (Lowest AIC = 21928.34)


The best model identified by stepwise selection includes:
adopted ~ sex + intake_condition + intake_type
This model reduces the AIC by 8,695 points compared to the null model, showing a large improved fit. 

Sex had the largest individual effect.
Intake condition captures medical and behavioral differences in animals.
Intake type reflects the context in which animals entered the shelter (for instance, foster, stray, TNR).

## Interpretting the Regression model and odds ratios
Model Overview
This logistic regression model estimates the log-odds of an animal being adopted based on: Animal type, Intake condition, Sex, Intake type

The baseline category (intercept) is: A male, unaltered dog with normal intake condition, from a foster intake type.

The model has:
Null deviance: 30,621 on 29,599 degrees of freedom
Residual deviance: 21,683 on 29,559 degrees of freedom
AIC: 21,765 — suggesting strong improvement over the null model.


Animal Type

- bird: Estimate = 1.044, p < 0.001

    - Birds are significantly more likely to be adopted than the baseline (dogs). Odds increase by exp(1.044) ≈ 2.84 times.
    
- cat: Estimate = 0.142, p = 0.0004

    - Cats also have higher odds than dogs; odds ratio ≈ 1.15.
    
- guinea pig: Estimate = 1.797, p < 2e-16

    - Very high adoption likelihood compared to dogs; odds ratio ≈ 6.03.
    
- livestock: Estimate = 2.574, p < 0.001

    - Strong effect, odds ratio ≈ 13.11.
    
- rabbit: Estimate = –0.579, p < 0.001

    - Rabbits are significantly less likely to be adopted than dogs; odds ratio ≈ 0.56.
    
- amphibian, wild, and reptile: Not statistically significant, most likely due to low sample sizes.


Intake Condition

- feral: Estimate = –2.693, p < 2e-16

    - Extremely reduced chance of adoption; odds ratio ≈ 0.07.
    
- fractious (aggressive): Estimate = –1.699, p < 2e-16

    - Significantly less likely to be adopted; odds ratio ≈ 0.18.
    
- behavior severe: Estimate = –1.189, p = 0.005

    - Odds of adoption ≈ 0.30.
    
- ill severe: Estimate = –0.875, p < 1e-9

    - Decreases odds to ≈ 0.42.
    
- injured severe: Estimate = –1.237, p < 2e-16

    - Reduces odds significantly to ≈ 0.29.
    
- under age/weight: Estimate = 0.220, p < 0.001

    - Slightly increases odds of adoption; odds ratio ≈ 1.25.
    
- behavior mild, ill mild, and injured mild: Not statistically significant.

Sex

- Neutered: Estimate = 2.407, p < 2e-16

    - Odds of adoption increase by exp(2.407) ≈ 11.10 times relative to intact males.
    
- Spayed: Estimate = 2.430, p < 2e-16

    - Odds increase by exp(2.430) ≈ 11.36.
    
- Unknown: Estimate = –1.871, p < 2e-16

    - Much lower odds of adoption; odds ratio ≈ 0.15.
    
- Female: Not statistically significant compared to males.
 
Intake Type

- trap, neuter, return (TNR): Estimate = –3.601, p = 0.023

    - Much lower likelihood of adoption (odds ratio ≈ 0.027), which makes sense since animals in this category are usually returned to their environment.
    
- Other intake types (e.g., stray, owner surrender) were not statistically significant, suggesting their effect is less pronounced or more variable in the dataset.

## Looking at each indivudal model

Animal Type

- Probability Plot
This shows the estimated probability of adoption by animal type.
    - Dogs and cats have moderate probabilities of adoption (~0.5–0.6).
    - Birds, rabbits, and guinea pigs show higher probabilities.
    - Livestock and reptiles also display high estimated probabilities, but these results may be due to low sample size or outlier cases (like adoptions through special programs).
    
- Residual Plot (AnimalResid.pdf)
The deviance residuals mostly fall within ±2, indicating good model fit across animal types.
    - A few larger residuals are visible for less common types like wild or reptile, suggesting predictions for those types might be less accurate (less data).
Conclusion: The model captures adoption trends well for major categories. Some rare animal types may need further data.

 Intake Condition 
 
- Probability Plot 
Shows a clear trend:
    - “Normal”, “under age/weight”, and “ill (mild)” animals have higher adoption probabilities.
    - Severely ill, injured, feral, or fractious animals show much lower probabilities.
    - Behavioral concerns—especially severe—lower adoption chances.
    
- Residual Plot 
Most values fall within ±2, suggesting overall good model fit.
    - Slight under- or overfitting is seen in categories with smaller sample sizes like “welfare seizures” and “i/i report”.
Conclusion: Intake condition is a strong predictor of adoption. Predictions for rare or ambiguous categories (like “i/i report”) might be less reliable.

 Intake Type 
 
- Probability Plot 
    - Animals from foster, owner surrender, and stray categories have higher adoption probabilities.
    - Trap-Neuter-Return (TNR) and wildlife have very low adoption probabilities—as expected since they are often returned, not adopted.
    
- Residual Plot 
Deviance residuals mostly within ±2 again, showing decent fit.
    - Some underprediction may occur for “stray” and “welfare seized,” possibly due to mixed adoption outcomes in those groups.
Conclusion: The model fits well, and the adoption probabilities align with operational shelter expectations. Intake type strongly influences adoption likelihood.

 Sex 
 
- Probability Plot
    - Spayed and neutered animals show significantly higher adoption probabilities.
    - Unaltered males and females are less likely to be adopted.
    - Unknown sex has the lowest probability, likely due to unclear health or behavior records.
    
- Residual Plot 
Residuals are mostly near zero, indicating excellent model fit.
Conclusion: Sterilization status is one of the most important predictors.

Overall Model Fit 
- This plot displays deviance residuals vs. fitted values.
    - Most residuals lie within the ±2 range across all predicted adoption probabilities.
    - There's no major curvature or funneling, which supports the model’s assumption of linearity on the logit scale and homoscedasticity.
Conclusion: The logistic model appears to be appropriate across the range of fitted probabilities. There are no signs of any large outliers.

# Conclusion

Overall, the best model we found that contained the lowest AIC value included four key predictions. These were animal type, intake condition, sex, and intake type. We found that sex was the strongest individual predictor for adoption likelihood. Sterilized animals were significantly more likely to be adopted than inteact males which was the baseline. The intake condition with the baseline as normally behaved reported that animals in poor health and behavioral issues were less likely to be adopted. Animals that were categorized as feral, fractious, or severe illness or injury had adoption odds 70-93% lower than healthy animals. Animal type is also a significant predictior. Our results showed that guinea pigs are 6 times more likely to be adopted than dogs and livestock was 13 times more likely to be adopted than dogs which was our baseline animal. Dogs however were the animal with the second highest population in the shelter. Lastly, types of stray, owner surrender, foster, and confinscate were not significantly significant predictors of adoption. TNR animals which stands for trapped, sterilized, and returned to their original location are not necessarily meant for adoption, however our model did show it was statistically significant. In the future, this is something to look more into specifically with TNR animals. Also in the future, it could be interesting to compare models when we group adoption vs. everything else as well as keeping the "everything else" categories separate. Overall, suggestions to the Long Beach Animal shelter would be to keep including variety in their animal selection because as we have seen in our results, it's not just perfect dogs and cats that get adopted. Every animal has a home.

### References

::: {#refs}
:::

# Appendix (Optional)

